Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-04-26 drafts/hybrid_zone.tex
minor
fixed fig 3 caption, more edits

The process of speciation commonly involves populations diverging with relatively little gene flow \citep{Coyne2004}.  
However, when formerly isolated populations come into contact before reproductive isolation is complete, some gene flow is possible.
Interbreeding and migration between such populations  
creates a gradient of alleles derived from the two populations across geographic space,
centered on their point of contact \citep[reviewed by][]{Barton1985}. 
If the populations are sufficiently diverged, 
this process leaves a distinct pattern of variation across the genome, 
in which long tracts of divergent haplotypes from each ancestral population  
are broken up by historical recombination events,
forming chromosomal {junctions} in ancestry \citep{Fisher1954, Chapman2002, baird2003distribution}. 
These patterns of correlated ancestry in admixed populations have been previously used to infer histories of hybridization and admixture 
\citep[e.g.][]{Gravel2012,Hellenthal2014,Harris2013,sedghifar2015spatial}.

@@

We model selection through a single, underdominant locus:
at this locus, there are two alleles, one corresponding to each of the ancestral populations.
Individuals who are heterozygous at this locus produce on average $1-s$ times as many offspring than either homozygote.
Although most selection in hybrid zones is likely more complex than this simple situation, 
previous work has shown that such models generate clines that are very similar to models of ecological selection or epistatic systems \citep{Kruuk1999,barton2000epistasis}, 
\yb{because selection on an allele depends on its marginal effect on fitness.  }
Exceptionally strong epistatic selection which acts more quickly than free recombination may therefore, yield results that differ slightly from ours. 
Likewise, while we do not model ecological selection, theoretical and simulation studies show that clines formed by underdominant selection and extrinsic selection are effectively indistinguishable without additional information \citep{may1975gene,Barton1993,Kruuk1999}, especially over short time scales \citep{Barton1979a}, \linelabel{ll:all_clines_are_the_same}
with their shape determined mostly by the mean fitness of the heterozygote \citep{Slatkin1973}.

@@

%%%%%%%%
\begin{figure}
\includegraphics[clip,width=\textwidth,trim=0cm 8cm 0cm 0.05cm]{figs/BM_schematic_new2.pdf}
\caption{
    \textbf{A schematic figure of the lineages along which a segment of genome has been inherited,}
    showing the location (horizontal axis) and time (vertical axis) at which the ancestors 
    of a hypothetical sampled haplotype lived.
    At the time of sampling ($\tau=0$), the segment contains the $B$ allele at the selected locus.
    Paths traced by lineages across space are depicted by blue and yellow lines. 
    Branching events occur when there was a recombination event within the sampled segment.
    The \alisa{rightmost} blue line represents the path of the selected locus. 
   % \plr{(All lines look the same width: no thick and thin.)}
    The colors of the other paths indicate the identity at the linked selected locus 
    (blue: linked to allele $B$; yellow: linked to allele $A$). 
    The center of the hybrid zone is at geographic position $x=0$; in the region $x<0$, ancestry $B$ is less common,
    and because of selection, the selected lineage spends little time there.
  %  Red dots represent ancestral recombination events that fall in the given segment of genome in an ancestor;
  %  these can cause the lineage to branch (lineages at linked sites are depicted as colored lines),
  %  or can switch lineages at linked sites between selected backgrounds 
  %  (yellow lineages are linked to the $A$ selected locus; blue are linked to $B$).
  %  Here the initial haplotype is broken up into 3 chunks over 100 generations,
  %  which switch between selected backgrounds once.
    Based on position of the lineages at $\tau=100$ generations ago,
    from left to right the segments are of ancestry $(A,B,B)$.
    % \plr{(Add blue/yellow bars at the bottom?)}
}\label{Fig:schematic}
\end{figure}

@@

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline,
affecting neighboring loci as well \citep{barton1979gene, barton1986barrier}.
First, we turn our attention to how the cline at the selected locus itself is formed.
%% we don't need to cite these here, since we cite Bazykin for the equation
% \alisa{Equations describing the forward-time establishment of the cline have been previously described \cite[e.g.][]{Slatkin1973,may1975gene,Nagylaki1975}}
Write $p(x,t)$ for the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and the $A$ allele is initially fixed on the left, so
$p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
We assume that dispersal is local and unbiased;
and that the mean squared displacement between parent and offspring
in the direction perpendicular to the cline
is $\sigma^2$.
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous,
and (iii) population density is large,
% and (iv) $s$ and $\sigma$ are small, \alisa{relative to $N$ and $\tau$ respectively?}
then as in \citet{Bazykin1969},
the commonly-used equation that $p$ approximately solves is
\begin{align} \label{eqn:cline_pde}
    \frac{d}{dt} p = \frac{\sigma^2}{2} \frac{d^2}{dx^2} p +  2 p (1-p)  \times s (\frac{p}{2} - 1), %YB Rearranged this equation to allow for cleaner description
    %    \frac{d}{dt} p = \frac{\sigma^2}{2} \frac{d}{dx^2} p + s p (1-p) (2p-1) ,
\end{align}
where $ \frac{d}{dt} p$ is the local rate of change of the frequency of $A$.
The first term, $\frac{\sigma^2}{2} \frac{d^2}{dx^2} p$, describes the net rate at which $A$ alleles arrive by local dispersal.
The second term % ,  $2 p (1-p)  s (p/2-1)$,  % only need this term in the sentence once?
describes the impact of selection on local allele frequency as the product of local genotypic variance, 
$2 p (1-p)$, and selection on allele $A$, $s (p/2-1)$. 
\plr{ I don't think 'genotypic variance' is the right interpretation?  Let me try again: 
The second term is the product of the local frequency of heterozygotes, $2p(1-p)$,
and the strength of selection on allele $A$, $s (p/2-1)$.  }
The term $s (p/2-1)$ describes underdominance -- $A$ is favored when common ($p>1/2$) and disfavored when rare ($p<1/2$).
%YB: Pardon the extreme changes here, I tried to make things more obvious to a biologist
%and a logistic term proportional to the difference in reproductive rate between $A$ and $a$ alleles
%due to local proportions of heterozygotes,
%which is $s(2p-1)$ (up to an error of size $s^2$).
The equation can be derived as in \citet{Nagylaki1975};
the assumption of locally random mating was shown to have only a small effect \citet{christiansen1995genotypic}.

@@

As noted by others \citep[e.g.][]{Slatkin1973,may1975gene}, \linelabel{ll:rescaling}
rescaling space and time by $\sigma/\sqrt{s}$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} 
results in a dimensionless equation,
implying that the cline establishes over a timescale of the order $1/s$. 
Since this is how long it takes diffusion at rate $\sigma^2$ to smooth across a region of width $\sigma/\sqrt{s}$,
this means we can think of
the cline as established by diffusion, 
despite being slowed down somewhat by selection against heterozygotes.

For simplicity we mostly work with one-dimensional clines,
but the mathematics applies as well to two-dimensional systems with some modifications:
for instance, if the hybrid zone is a straight line, 
the description above applies to motion of a lineage transverse to the hybrid zone.  \revpoint{AE}{4}
In this case, $d^2/dx^2$ should be replaced by the Laplacian operator,
and $d/dx$ by the gradient.
In fact, if the landscape is heterogeneous or dispersal is non-Gaussian,
the theory still holds replacing the Laplacian by the appropriate operator.
Note, however, that with low population densities, 
drift can have a strong effect that differs between one and two dimensional systems \citep{cox1995hybrid,barton2010phylogeography,durrett2007width}.

@@

Making the same assumptions as for equation \eqref{eqn:cline_pde}, the derivation in Appendix \ref{apx:lineage_derivation} shows 
that the lineage of an $A$ allele at position $x$ moves as Brownian motion driven by dispersal
with mean displacement proportional to the derivative of $\log p$.
This has the effect of pulling $A$ lineages towards regions where they are more common, i.e., the left side of the range.   
% with speed $\sigma$
% with bias $\sigma^2 \grad \log p(x,t)$, where $\grad$ is the gradient
(the analogous physical system is a particle that moves at speed $\sigma^2$ in the potential $-\log p$). 
%\alisa{What does potential here mean?}
%\plr{(It means like a gravity well.  Clear now that it says ``particle''?)}
To see why this is true, note that the probability that the parent of an $A$ allele found at $x$
lived at position $x-r$ is proportional to $\P\{R=r\} p(x-r)$, where $R$ is the random dispersal distance;
and so the mean displacement from offspring to parent is $\E[R p(x-R)]/\E[p(x-R)]$.  \revpoint{1}{9}
% $$ = \E[ -R ] - \E[R]^2 p'(x)/p(x) + \E[R^2] p'(x)/p(x) + O(R^3) $$
Using the fact that $\E[R]=0$ and $\E[R^2]=\sigma^2$ 
%\alisa{(perhaps a brief explanation of how this works; in particular I wasn't sure why were were taking the ratio of two expectations...)}
%\plr{(That's what the rest of the sentence is? Work it out, and if you still think it needs more detail, add it?)}
and expanding $p()$ to first order about $x$ shows that this mean displacement is approximately $\sigma^2 p'(x)/p(x)$,
which is $\sigma^2$ multiplied by the gradient of $\log p(x)$.
This description holds even when the frequency profile of the $A$ allele changes with time (replacing $p(x)$ by $p(x,t)$).

@@

This description gives more information than the steady-state cline,
which depends only on $\sigma/\sqrt{s}$.
Here we see that lineages with $\sigma=10$ and $s=.16$ move much faster
than lineages with $\sigma=1$ and $s=.0016$,
reflecting strong differences in selection against heterozygotes,
even though the stable clines have the same form, and therefore have additional information about the hybrid zone.
Even though a rescaling of space and time as described above can make the models equivalent,
this difference in lineage speed can be seen through the action of recombination, which we explore next.
\revpoint{AE}{3}

@@

\yb{Suppose we sample an allele today $r$ Morgans from the selected site, and  
follow its lineage back through time, using $\tau$ to denote ``generations ago''
(reserving $t$ for time measured in the forward direction).
If $X_\tau$ is the geographic location of its ancestor $\tau$ generations ago,
then we say that $X$ moves as a diffusion pushed by either $\log(p)$ or $\log(1-p)$ 
(as described for a selected allele in equation \eqref{eqn:cline_pde}).
Following this lineage back in time, the identity of the selected allele that ancestor carried at time $\tau$, $Z_\tau$, 
jumps between $A$ and $B$ with recombination events between the neutral and selected locus (with frequency $r$), that result in a change in the  selected background.    
Thus, by the assumption of locally random assortment of alleles, $Z$ shifts from $A$ to $B$  at rate $r (1-p)$, and $B$ to $A$ at rate $r p$ 
(see Appendix \ref{apx:lineage_derivation} for a more precise description).}

We can describe this Markov process formally in It\^o notation:
with $B_\tau$ a standard Brownian motion,
$T_B(\tau)$ the most recent time before $\tau$ that $Z_\tau = B$, and likewise for $T_A$,
\begin{align}
    \begin{aligned} \label{eqn:lineage_motion}
        dX_\tau &= \sigma dB_\tau + \begin{cases}
            \sigma^2 \frac{d}{dx} \log(p(X_\tau,\tau)) d\tau \qquad & \text{if } Z_\tau = A \\
             \sigma^2 \frac{d}{dx} \log(1-p(X_\tau,\tau)) d\tau \qquad & \text{if } Z_\tau = B 
        \end{cases} \\
        \P\{ Z_{\tau+\epsilon} = B \given Z_\tau = A, \; X_\tau=x \} &= \epsilon r (1-p(x,\tau)) + O(\epsilon^2) \\
        \P\{ Z_{\tau+\epsilon} = A \given Z_\tau = B, \; X_\tau=x \} &= \epsilon r p(x,\tau) + O(\epsilon^2)  .
        % \P\{ T_B(\tau) &> \tau+u \given Z_\tau = A \} = \exp\left( - r \int_\tau^{\tau+u} (1-p(X_s,s)) ds \right) \\
        % \P\{ T_A(\tau) &> \tau+u \given Z_\tau = B \} = \exp\left( - r \int_\tau^{\tau+u} p(X_s,s) ds \right) .
        % \E[ T_B(\tau) &> \tau+u \given Z_\tau = A ] = r \int_\tau^{\tau+u} (1-p(X_s,s)) ds \\
        % \E[ T_A(\tau) &> \tau+u \given Z_\tau = B ] = r \int_\tau^{\tau+u} p(X_s,s) ds .
    \end{aligned}
\end{align}
In the first expression, giving the distribution of how the lineage location changes,
the first term ($\sigma dB_\tau$) is Brownian noise driven by dispersal,
and the second is the mean displacement, which moves the lineage ``downhill'' towards its selected allele's ancestral range on either $-\log p$ or $-\log (1-p)$,
depending on which selected allele the lineage is linked to.
(In two dimensions, $d/dx$ is replaced by the gradient.)
\plr{
The second expression says that the probability a lineage on the $A$ background at $x$
recombines onto the $B$ background
is equal to the recombination rate, $r$, multiplied by the local proportion of $B$ alleles, $(1-p(x,\tau))$,
per generation.
}
% \plr{
% The second expression gives the probability that a lineage on the $A$ background at time $\tau$
% does not switch to the $B$ background across $u$ generations,
% i.e., the probability that there was no recombination between the lineage and the selected site in those $u$ generations.
% Since the rate of recombination between the two is $r$,
% and the probability the lineage is in a heterozygote at time $s$ 
% is equal to the local proportion of $B$ alleles, $1-p(X_s,s)$,
% the mean number of recombinations in heterozygotes across that time period is $r \int_\tau^{\tau+u} (1-p(X_s,s)) ds$;
% assuming recombinations are Poisson, the probability of no such recombinations is the expression given.
% The third expression is the same, except for switching $B$ to $A$.
% }
% gives the rate at which a lineage switches from an $A$ to $B$ background as a result of recombination,
% which occurs at rate $r$ multiplied by the local proportion of $B$ alleles, which is $(1-p)$.
% The third expression is the corresponding rate for switching from $B$ to $A$.

@@

\paragraph{Linked clines}
We can use this diffusion model for lineages to find expected clines in ancestry,
i.e., the expected proportion of individuals who inherit from species $A$,
as a function of space, time, and position on the genome.
Precisely, we need the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$,
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
We denote this probability $q_z(x,t,r)$.
In the notation above,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A \given X_0 = x, \; Z_0 = z\} .
\end{align}
(Recall that $Z$ depends implicitly on $r$.) \revpoint{1}{7}
The description above implies that $q_z$ solves the following Kolmogorov backward equation:
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \frac{d}{dt} q_A(x,t,r) 
            &= 
                \frac{\sigma^2}{2} \frac{d^2}{dx^2} q_A(x,t,r) 
                + \sigma^2 \frac{d}{dx} \log(p(x,t)) \cdot \frac{d}{dx} q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
        \frac{d}{dt} q_B(x,t,r) 
            &= 
                \frac{\sigma^2}{2} \frac{d^2}{dx^2} q_B(x,t,r)
                + \sigma^2 \frac{d}{dx} \log(1-p(x,t)) \cdot \frac{d}{dx} q_B(x,t,r) 
            \\ &\qquad {} + 
            r p(x,t) (q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} 
\end{align}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$.
% and where $\grad$ is the gradient.
The three terms in these equations come from Brownian movement of a linages 
(i.e., the smoothing action of local dispersal),
the tendency of a lineage to inherit from regions where its type is more common 
(i.e., the net flux induced by reduced fitness in the hybrid zone),
and recombination between selected backgrounds, respectively. \alisa{I think the first and second terms may have been switched??}
\yb{We note that because recombination occurs within a deme, the recombination terms in equation \eqref{eqn:q_pde} are quite standard \citep{HartlClark}.} \revpoint{1}{4}
%\alisa{this is fwd-in-time notation. should we change it to bwd-in-time?}
%\plr{This equation should be forwards in time, since it's the Kolmogorov backward equation
%applied to a backwards-time process!}

We will have more use for the differential operators on the right-hand sides of these equations,
so define these as
$G^A = \sigma^2 \frac{d}{dx} \log(p(x,t)) \cdot \frac{d}{dx} + (\sigma^2/2) \frac{d}{dx^2}$
and
$G^B = \sigma^2 \frac{d}{dx} \log(1-p(x,t)) \cdot \frac{d}{dx} + (\sigma^2/2) \frac{d}{dx^2}$,
so that equation \eqref{eqn:q_pde} can be written more compactly as
\begin{align*}
    \frac{d}{dt} q_A &= G^A q_A + r (1-p) (q_B-q_A) \\
    \frac{d}{dt} q_B &= G^B q_B + r p (q_A-q_B) .
\end{align*}
In technical terms, $G^A$ and $G^B$ are the generators of the diffusions of lineages of selected alleles of ancestry $A$ and $B$, respectively;
informally, they encode the stochastic motion of lineages at the selected loci.

@@

\plr{Below we use $l_B$ to mean the normalized version also.  I am defining consistent notation, $\bar l_B$.}

@@

\paragraph{The nature of selection}
Although our assumed scenario of selection against heterozygotes at a single locus is uncommon in nature, 
we believe that this is unlikely to drastically influence our findings.
As discussed in the Introduction,
previous studies have demonstrated that models of underdominant selection 
share similar properties with more realistic models, e.g.,
selection on an ecological gradient or multilocus, epistatic selection
\citep{Slatkin1973,Barton1993,Kruuk1999}.
\revpoint{1}{1}

>>>>

The process of speciation commonly involves populations diverging with relatively little gene flow \citep{Coyne2004}.  
However, when formerly isolated populations come into contact before reproductive isolation is complete, some gene flow is possible.
Interbreeding and migration between such populations  
creates a gradient of alleles derived from the two populations across geographic space,
centered on their point of contact \citep[reviewed by][]{Barton1985}. 
If the populations are sufficiently diverged, 
this process leaves a distinct pattern of variation across the genome, 
in which long tracts of divergent haplotypes from each ancestral population  
are broken up by historical recombination events,
forming chromosomal {junctions} in ancestry \citep{Fisher1954, Chapman2002, baird2003distribution}. 
These patterns of correlated ancestry in admixed populations have been previously used to infer histories of hybridization and admixture 
\citep[e.g.,][]{Gravel2012,Hellenthal2014,Harris2013,sedghifar2015spatial}.

@@

We model selection through a single, underdominant locus:
at this locus, there are two alleles, one corresponding to each of the ancestral populations.
Individuals who are heterozygous at this locus produce on average $1-s$ times as many offspring than either homozygote.
Although most selection in hybrid zones is likely more complex than this simple situation, 
previous work has shown that such models generate clines that are very similar to models of ecological selection or epistatic systems \citep{Kruuk1999,barton2000epistasis}, 
because selection on an allele depends on its marginal effect on fitness.
(More discussion of this point in the Discussion.) \linelabel{ll:all_clines_are_the_same}

@@

%%%%%%%%
\begin{figure}
\includegraphics[clip,width=\textwidth,trim=0cm 8cm 0cm 0.05cm]{figs/BM_schematic_new2.pdf}
\caption{
    \textbf{A schematic figure of the lineages along which a segment of genome has been inherited,}
    showing the location (horizontal axis) and time (vertical axis) at which the ancestors 
    of a hypothetical sampled haplotype lived.
    At the time of sampling ($\tau=0$), the segment contains the $B$ allele at the selected locus.
    Paths traced by lineages across space are depicted by blue and yellow lines. 
    Branching events occur when there was a recombination event within the sampled segment.
    The rightmost blue line represents the path of the selected locus. 
    The colors of the other paths indicate the identity at the linked selected locus 
    (blue: linked to allele $B$; yellow: linked to allele $A$). 
    The center of the hybrid zone is at geographic position $x=0$; in the region $x<0$, ancestry $B$ is less common,
    and because of selection, the selected lineage spends little time there.
  %  Red dots represent ancestral recombination events that fall in the given segment of genome in an ancestor;
  %  these can cause the lineage to branch (lineages at linked sites are depicted as colored lines),
  %  or can switch lineages at linked sites between selected backgrounds 
  %  (yellow lineages are linked to the $A$ selected locus; blue are linked to $B$).
  %  Here the initial haplotype is broken up into 3 chunks over 100 generations,
  %  which switch between selected backgrounds once.
    Based on position of the lineages at $\tau=100$ generations ago,
    from left to right the segments are of ancestry $(A,B,B)$.
    % \plr{(Add blue/yellow bars at the bottom?)}
}\label{Fig:schematic}
\end{figure}

@@

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline,
affecting neighboring loci as well \citep{barton1979gene, barton1986barrier}.
First, we turn our attention to how the cline at the selected locus itself is formed.
%% we don't need to cite these here, since we cite Bazykin for the equation
% \alisa{Equations describing the forward-time establishment of the cline have been previously described \cite[e.g.,][]{Slatkin1973,may1975gene,Nagylaki1975}}
Write $p(x,t)$ for the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and the $A$ allele is initially fixed on the left, so
$p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
We assume that dispersal is local and unbiased;
and that the mean squared displacement between parent and offspring
in the direction perpendicular to the cline
is $\sigma^2$.
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous,
and (iii) population density is large,
% and (iv) $s$ and $\sigma$ are small, \alisa{relative to $N$ and $\tau$ respectively?}
then as in \citet{Bazykin1969},
the commonly-used equation that $p$ approximately solves is
\begin{align} \label{eqn:cline_pde}
    \frac{d}{dt} p = \frac{\sigma^2}{2} \frac{d^2}{dx^2} p +  2 p (1-p)  \times s (\frac{p}{2} - 1).
    %    \frac{d}{dt} p = \frac{\sigma^2}{2} \frac{d}{dx^2} p + s p (1-p) (2p-1) ,
\end{align}
Here the frequency $p=p(x,t)$ varies across time and space,
and so $ \frac{d}{dt} p$ is the local rate of change of the frequency of $A$.
The first term, $\frac{\sigma^2}{2} \frac{d^2}{dx^2} p$, describes the net rate at which $A$ alleles arrive by local dispersal.
The second term % ,  $2 p (1-p)  s (p/2-1)$,  % only need this term in the sentence once?
describes the impact of selection on local allele frequency as the product of local genotypic variance, $2p(1-p)$
(equivalently, the local frequency of heterozygotes), 
and selection on allele $A$, $s (p/2-1)$. 
The term $s (p/2-1)$ describes underdominance -- $A$ is favored when common ($p>1/2$) and disfavored when rare ($p<1/2$).
%YB: Pardon the extreme changes here, I tried to make things more obvious to a biologist
%and a logistic term proportional to the difference in reproductive rate between $A$ and $a$ alleles
%due to local proportions of heterozygotes,
%which is $s(2p-1)$ (up to an error of size $s^2$).
The equation can be derived as in \citet{Nagylaki1975};
the assumption of locally random mating was shown to have only a small effect \citet{christiansen1995genotypic}.

@@

As noted by others \citep[e.g.,][]{Slatkin1973,may1975gene}, \linelabel{ll:rescaling}
rescaling space and time by $\sigma/\sqrt{s}$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} 
results in a dimensionless equation,
implying that the cline establishes over a timescale of the order $1/s$. 
Since this is how long it takes diffusion at rate $\sigma^2$ to smooth across a region of width $\sigma/\sqrt{s}$,
this means we can think of
the cline as established by diffusion, 
despite being slowed down somewhat by selection against heterozygotes.

For simplicity we mostly work with one-dimensional clines,
but the mathematics applies as well to two-dimensional systems with some modifications:
for instance, if the hybrid zone is a straight line, 
the description above applies to motion of a lineage transverse to the hybrid zone.  \revpoint{AE}{4}
In this case, $d^2/dx^2$ should be replaced by the Laplacian operator.
% and $d/dx$ by the gradient.
In fact, if the landscape is heterogeneous or dispersal is non-Gaussian,
the theory still holds replacing the Laplacian by the appropriate operator.
Note, however, that with low population densities, 
drift can have a strong effect that differs between one and two dimensional systems \citep{cox1995hybrid,barton2010phylogeography,durrett2007width}.

@@

Making the same assumptions as for equation \eqref{eqn:cline_pde}, the derivation in Appendix \ref{apx:lineage_derivation} shows 
that the lineage of an $A$ allele at position $x$ moves as Brownian motion driven by dispersal
with mean displacement proportional to the spatial derivative of $\log p$.
This has the effect of pulling $A$ lineages towards regions where they are more common, i.e., the left side of the range.   
% with speed $\sigma$
% with bias $\sigma^2 \grad \log p(x,t)$, where $\grad$ is the gradient
(The analogous physical system is a particle that moves at speed $\sigma^2$ in the potential $-\log p$). 
%\alisa{What does potential here mean?}
%\plr{(It means like a gravity well.  Clear now that it says ``particle''?)}
To see why this is true, note that the probability that the parent of an $A$ allele found at $x$
lived at position $x-r$ is proportional to $\P\{R=r\} p(x-r)$, where $R$ is the random dispersal distance;
and so the mean displacement from offspring to parent is $\E[R p(x-R)]/\E[p(x-R)]$.  \revpoint{1}{9}
% $$ = \E[ -R ] - \E[R]^2 p'(x)/p(x) + \E[R^2] p'(x)/p(x) + O(R^3) $$
Using the fact that $\E[R]=0$ and $\E[R^2]=\sigma^2$ 
%\alisa{(perhaps a brief explanation of how this works; in particular I wasn't sure why were were taking the ratio of two expectations...)}
%\plr{(That's what the rest of the sentence is? Work it out, and if you still think it needs more detail, add it?)}
and expanding $p()$ to first order about $x$ shows that this mean displacement is approximately $\sigma^2 p'(x)/p(x)$,
which is $\sigma^2$ multiplied by the gradient of $\log p(x)$.
This description holds even when the frequency profile of the $A$ allele changes with time (replacing $p(x)$ by $p(x,t)$).

@@

This description gives more information than the steady-state cline,
which depends only on $\sigma/\sqrt{s}$.
Here we see that lineages with $\sigma=10$ and $s=.16$ move much faster
than lineages with $\sigma=1$ and $s=.0016$,
reflecting strong differences in selection against heterozygotes,
even though the stable clines have the same form.
Even though a rescaling of space and time as described above can make the models equivalent,
this difference in lineage speed can be seen through the action of recombination, which we explore next.
\revpoint{AE}{3}

@@

Suppose we sample an allele today $r$ Morgans from the selected site, and  
follow its lineage back through time, using $\tau$ to denote ``generations ago''
(reserving $t$ for time measured in the forward direction).
If $X_\tau$ is the geographic location of its ancestor $\tau$ generations ago,
then we say that $X$ moves as a diffusion pushed by either $\log(p)$ or $\log(1-p)$ 
(as described for a selected allele in the previous section).
Following this lineage back in time, the identity of the selected allele that ancestor carried at time $\tau$, $Z_\tau$, 
jumps between $A$ and $B$ with each recombination event that occurs between the focal locus and the selected locus (with frequency $r$), 
that results in a change in the  selected background.    
Thus, by the assumption of locally random assortment of alleles, $Z$ shifts from $A$ to $B$  at rate $r (1-p)$, and $B$ to $A$ at rate $r p$ 
(see Appendix \ref{apx:lineage_derivation} for a more precise description).

We can describe this Markov process formally in It\^o notation:
with $B_\tau$ a standard Brownian motion,
$T_B(\tau)$ the most recent time before $\tau$ that $Z_\tau = B$, and likewise for $T_A(\tau)$,
\begin{align}
    \begin{aligned} \label{eqn:lineage_motion}
        dX_\tau &= \sigma dB_\tau + \begin{cases}
            \sigma^2 \frac{d}{dx} \log(p(X_\tau,\tau)) d\tau \qquad & \text{if } Z_\tau = A \\
             \sigma^2 \frac{d}{dx} \log(1-p(X_\tau,\tau)) d\tau \qquad & \text{if } Z_\tau = B 
        \end{cases} \\
        \P\{ Z_{\tau+\epsilon} = B &\given Z_\tau = A, \; X_\tau=x \} = \epsilon r (1-p(x,\tau)) + O(\epsilon^2) \\
        \P\{ Z_{\tau+\epsilon} = A &\given Z_\tau = B, \; X_\tau=x \} = \epsilon r p(x,\tau) + O(\epsilon^2)  .
        % \P\{ T_B(\tau) &> \tau+u \given Z_\tau = A \} = \exp\left( - r \int_\tau^{\tau+u} (1-p(X_s,s)) ds \right) \\
        % \P\{ T_A(\tau) &> \tau+u \given Z_\tau = B \} = \exp\left( - r \int_\tau^{\tau+u} p(X_s,s) ds \right) .
        % \E[ T_B(\tau) &> \tau+u \given Z_\tau = A ] = r \int_\tau^{\tau+u} (1-p(X_s,s)) ds \\
        % \E[ T_A(\tau) &> \tau+u \given Z_\tau = B ] = r \int_\tau^{\tau+u} p(X_s,s) ds .
    \end{aligned}
\end{align}
In the first expression, giving the distribution of how the lineage location changes,
the first term ($\sigma dB_\tau$) is Brownian noise driven by dispersal,
and the second is the mean displacement, which moves the lineage ``downhill'' towards its selected allele's ancestral range on either $-\log p$ or $-\log (1-p)$,
depending on which selected allele the lineage is linked to.
(In two dimensions, $d/dx$ is replaced by the gradient.)
The second expression says that the probability a lineage on the $A$ background at $x$
recombines onto the $B$ background
is equal to the recombination rate, $r$, multiplied by the local proportion of $B$ alleles, $(1-p(x,\tau))$,
per generation.
% \plr{
% The second expression gives the probability that a lineage on the $A$ background at time $\tau$
% does not switch to the $B$ background across $u$ generations,
% i.e., the probability that there was no recombination between the lineage and the selected site in those $u$ generations.
% Since the rate of recombination between the two is $r$,
% and the probability the lineage is in a heterozygote at time $s$ 
% is equal to the local proportion of $B$ alleles, $1-p(X_s,s)$,
% the mean number of recombinations in heterozygotes across that time period is $r \int_\tau^{\tau+u} (1-p(X_s,s)) ds$;
% assuming recombinations are Poisson, the probability of no such recombinations is the expression given.
% The third expression is the same, except for switching $B$ to $A$.
% }
% gives the rate at which a lineage switches from an $A$ to $B$ background as a result of recombination,
% which occurs at rate $r$ multiplied by the local proportion of $B$ alleles, which is $(1-p)$.
% The third expression is the corresponding rate for switching from $B$ to $A$.

@@

\paragraph{Linked clines}
We can use this diffusion model for lineages to find expected clines in ancestry,
i.e., the expected proportion of individuals who inherit from species $A$,
as a function of space, time, and position on the genome.
Precisely, we need the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$,
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
We denote this probability $q_z(x,t,r)$.
In the notation above,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A \given X_0 = x, \; Z_0 = z\} .
\end{align}
(Recall that $Z$ depends implicitly on $r$.) \revpoint{1}{7}
The description above implies that $q_z$ solves the following Kolmogorov backward equation:
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \frac{d}{dt} q_A(x,t,r) 
            &= 
                \frac{\sigma^2}{2} \frac{d^2}{dx^2} q_A(x,t,r) 
                + \sigma^2 \frac{d}{dx} \log(p(x,t)) \cdot \frac{d}{dx} q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
        \frac{d}{dt} q_B(x,t,r) 
            &= 
                \frac{\sigma^2}{2} \frac{d^2}{dx^2} q_B(x,t,r)
                + \sigma^2 \frac{d}{dx} \log(1-p(x,t)) \cdot \frac{d}{dx} q_B(x,t,r) 
            \\ &\qquad {} + 
            r p(x,t) (q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} 
\end{align}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$.
% and where $\grad$ is the gradient.
The three terms in these equations come from Brownian movement of a linages 
(i.e., the smoothing action of local dispersal),
the tendency of a lineage to inherit from regions where its type is more common 
(i.e., the net flux induced by reduced fitness in the hybrid zone),
and recombination between selected backgrounds, respectively.
We note that because recombination occurs within a deme, the recombination terms in equation \eqref{eqn:q_pde} are quite standard \citep{HartlClark}. \revpoint{1}{4}
%\alisa{this is fwd-in-time notation. should we change it to bwd-in-time?}
%\plr{This equation should be forwards in time, since it's the Kolmogorov backward equation
%applied to a backwards-time process!}

We will have more use for the differential operators on the right-hand sides of these equations,
so define these as
$
G^A = 
(\sigma^2/2) \frac{d}{dx^2}
+ 
\sigma^2 \frac{d}{dx} \log(p(x,t)) \cdot \frac{d}{dx} 
$
and
$
G^B = 
(\sigma^2/2) \frac{d}{dx^2}
+ 
\sigma^2 \frac{d}{dx} \log(1-p(x,t)) \cdot \frac{d}{dx} 
$
,
so that equation \eqref{eqn:q_pde} can be written more compactly as
\begin{align*}
    \frac{d}{dt} q_A &= G^A q_A + r (1-p) (q_B-q_A) \\
    \frac{d}{dt} q_B &= G^B q_B + r p (q_A-q_B) .
\end{align*}
In technical terms, $G^A$ and $G^B$ are the generators of the diffusions of lineages of selected alleles of ancestry $A$ and $B$, respectively;
informally, they encode the stochastic motion of lineages at the selected loci.

@@

@@

\paragraph{The nature of selection}
Although our assumed scenario of selection against heterozygotes at a single locus is uncommon in nature, 
theoretical and simulation studies show that clines formed by underdominant selection and extrinsic selection 
are effectively indistinguishable without additional information \citep{may1975gene,Barton1993,Kruuk1999}, 
especially over short time scales \citep{Barton1979a},
and their shape is determined mostly by the mean fitness of the heterozygote \citep{Slatkin1973}. 
\revpoint{1}{1}
